A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-02-09 sister/sister.tex
more tuning

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) 
\kern2.0em %% TYPESETTING
\Val \ec{\yasx}  \,=\,
\MU {\Varec{\yasx}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{\yasx}} ) )
%\kern0.25em %% TYPESETTING
% &
\\
\Val \ec{\const{S}(y)} & \,=\, 
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{\yasx}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{\yasx} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{\yasx}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{\yasx}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again, % to XXX the equivalence classes,
yielding the equivalence classes
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{\yasx} & \,=\, \MU {\Varec{\yasx}}.\; \const{S}({\Varec{\yasx}})
\end{align*}
Since $\Val \ec{x} \aequiv \Val \ec{\yasx}$,
%for the renaming
%$\{ {\Varec{\yasx}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{\yasx}} \}$,
\rn{Unique} can finally be applied to derive $x \teq y$.
\xend
\end{examplex}

@@

If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
$t$'s equivalence class must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
Another rule, \rn{Single}, focuses on the degenerate case where two
terms %from $\tEc$
are of a \emph{singleton} type (one for which there exists only one value),
and are therefore equal.
Notice that corecursive singleton types may have infinite values. A simple example is
\vthinspace$\keyw{codata}~\ty{a} = \const{A}(\ty{a})$, which is corecursive
and yet has a cardinality of one;
its unique value is $\MU a.\; \const{A}(a)$.

@@

Both
\rn{Split}'s finiteness assumption %on the datatype
and \rn{Single}'s singleton
constraint %on the codatatype
can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types.
%This step relies on the assumption that uninterpreted types are infinite.

@@

\begin{theorem}[Termination]%
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theorem}

\begin{theorem}[Refutation Soundness]%
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is %$\thD$-
unsatisfiable.
\end{theorem}

\begin{theorem}[Solution Soundness]%
\label{thm:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is %$\thD$-
satisfiable.
\end{theorem}

By Theorems~\ref{thm:t},~\ref{thm:rs}, and~\ref{thm:ss}, the
calculus is sound and complete for the universal theory of (co)datatypes. We may
rightly call it a decision procedure for that theory.
The proof of solution soundness is constructive in that it provides a
method for constructing a model for a saturated configuration.
%AJR : if necessary...
%To give an intuition how this model is constructed,
%we consider the mapping $\Val$ from Step 2 of the procedure,
%and apply additional substitutions over the free variables in its range
%to obtain a mapping, call it $\ValC$, 
%from equivalence classes to closed $\MU$-terms.
%Then, for each equivalence class $\ec{t}$ in the domain of $\ValC$,
%we interpret each variable in $\ec{t}$ as $\ValC \ec{t}$.
%The properties of a saturated configuration ensure
%that $\ValC$ can be constructed such that all constraints in our input are satisfied.

@@

\newcommand\captioni{\,Number of solved goals with $n$~lemmas per goal}
\newcommand\captionii{\,Number of solved goals with 256~lemmas per goal}

@@

\begin{table}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{$n = 16$\,} & \multicolumn{2}{@{}c@{\;}}{$n = 256$\;} \\%[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{\,Z3}
\MIDRULE
No (co)datatypes
 &  1099  &  1097  &  2209  &  1911 \\%[\jot]
Datatypes without \rn{Acyclic}
 &  1116  &  -- &  2211  &  -- \\
Full datatypes
 &  1120  &  1121  &  2211  &  1901 \\%[\jot]
Codatatypes without \rn{Unique}\!\!
 &  1132  &  --  &  2208  &  -- \\
Full codatatypes
 &  1137  &  -- &  \bfseries 2220  &  -- \\%[\jot]
Full (co)datatypes
 &  \bfseries 1157  &  -- &  2219  &  -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:bench}
\end{table}

>>>>

\theoremstyle{plain}
\newtheorem*{theoremi}{Theorem 1 (Termination)}
\newtheorem*{theoremii}{Theorem 2 (Refutation Soundness)}
\newtheorem*{theoremiii}{Theorem 3 (Solution Soundness)}

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) 
\kern2.0em %% TYPESETTING
\Val \ec{\yasx}  \,=\,
\MU {\Varec{\yasx}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{\yasx}} ) )
%\kern0.25em %% TYPESETTING
% &
\\[-.5\jot]
\Val \ec{\const{S}(y)} & \,=\, 
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{\yasx}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{\yasx} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{\yasx}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{\yasx}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again, % to XXX the equivalence classes,
yielding the equivalence classes
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{\yasx} & \,=\, \MU {\Varec{\yasx}}.\; \const{S}({\Varec{\yasx}})
\end{align*}
Since $\Val \ec{x} \aequiv \Val \ec{\yasx}$,
%for the renaming
%$\{ {\Varec{\yasx}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{\yasx}} \}$,
\rn{Unique} can finally derive $x \teq y$.
\xend
\end{examplex}

@@

If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
$t$'s equivalence class must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
Another rule, \rn{Single}, focuses on the degenerate case where two
terms %from $\tEc$
are of a \emph{singleton} type (one for which there exists only one value),
and are therefore equal.
Notice that corecursive singleton types may have infinite values. A simple example is
\vthinspace$\keyw{codata}~\ty{a} = \const{A}(\ty{a})$, which is corecursive
and yet has a cardinality of one;
its unique value is $\MU a.\; \const{A}(a)$.
Both
\rn{Split}'s finiteness assumption %on the datatype
and \rn{Single}'s singleton
constraint %on the codatatype
can be evaluated statically
based on a recursive computation of the cardinalities of the
constructors' argument types.
%This step relies on the assumption that uninterpreted types are infinite.

@@

@@

\begin{theoremi}
\label{thm:t}%
\afterDot
All derivation trees are finite.
\end{theoremi}

\begin{theoremii}
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is %$\thD$-
unsatisfiable.
\end{theoremii}

\begin{theoremiii}
\label{thm:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is %$\thD$-
satisfiable.
\end{theoremiii}

\looseness=-1
By Theorems~1, 2, and 3, the
calculus is sound and complete for the universal theory of (co)datatypes. We may
rightly call it a decision procedure for that theory.
The proof of solution soundness is constructive in that it provides a
method for constructing a model for a saturated configuration.

%AJR : if necessary...
%To give an intuition how this model is constructed,
%we consider the mapping $\Val$ from Step 2 of the procedure,
%and apply additional substitutions over the free variables in its range
%to obtain a mapping, call it $\ValC$, 
%from equivalence classes to closed $\MU$-terms.
%Then, for each equivalence class $\ec{t}$ in the domain of $\ValC$,
%we interpret each variable in $\ec{t}$ as $\ValC \ec{t}$.
%The properties of a saturated configuration ensure
%that $\ValC$ can be constructed such that all constraints in our input are satisfied.

@@

\newcommand\captioni{\,Number of solved goals with $n$~lemmas per goal}

@@

\begin{table}[h!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{$n = 16$\,} & \multicolumn{2}{@{}c@{\;}}{$n = 256$\;} \\%[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{\,Z3}
\MIDRULE
No (co)datatypes
 &  1099  &  1097  &  2209  &  1911 \\%[\jot]
Datatypes without \rn{Acyclic}
 &  1116  &  -- &  2211  &  -- \\
Full datatypes
 &  1120  &  1121  &  2211  &  1901 \\%[\jot]
Codatatypes without \rn{Unique}\!\!
 &  1132  &  --  &  2208  &  -- \\
Full codatatypes
 &  1137  &  -- &  \bfseries 2220  &  -- \\%[\jot]
Full (co)datatypes
 &  \bfseries 1157  &  -- &  2219  &  -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:bench}
\end{table}
